---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Interactive Tools

    Here are some tools to help with interactive viewing of data. We use plotly for visualizing data in gpplot, and leaflet and mapview for visualizing spatial data. 

## Plotly - Interactive plotting

Load packages.
```{r , warning = FALSE, message = FALSE}
library(ggplot2)
library(lubridate)
library(plotly)
```

We are using the iris dataset that comes with R.
```{r}
head(iris)
```

Make a plot. 
```{r, warning = FALSE, message = FALSE}
iris1 <- ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) + geom_point()
```

Use plotly to make plot interactive. 
You can draw a box around an area to zoom in, double click on plot to zoom back out. 
If you click on virginica in the legend it will remove that from the plot.
```{r}
ggplotly(iris1)
```

You might want to change what shows up when you hover over the point. You can do this by adding the "label" parameter to your aes() call.
```{r}
iris2 <- ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species, label = Petal.Length)) + geom_point()
ggplotly(iris2)
```

If you want even more customization of what shows up, you can use the "text" parameter in your aes() call. You can then customize the combination of what shows up: x, y, text.

Note '<br>' inserts a line break into your popup. 
```{r, warning = FALSE, message = FALSE}
iris3 <- ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width,
                    text = paste('Name of species: ', Species,
                                 '<br>Petal length: ', Petal.Length)))+
  geom_point()

# customize whether you want to display x, y, and text, or just text
ggplotly(iris3)
ggplotly(iris3, tooltip = "text") 
ggplotly(iris3, tooltip = c("x", "text"))
```

If you don't want the popup at all, you can also get rid of it. 
```{r}
# Use tooltip = NULL to get rid of all tooltip 
ggplotly(iris3, tooltip = NULL)
```

Resources: 
* https://plotly.com/r/hover-text-and-formatting/
* https://stackoverflow.com/questions/36325154/how-to-choose-variable-to-display-in-tooltip-when-using-ggplotly


## Leaflet

```{r, warning = FALSE, message = FALSE}
library(readr)
library(leaflet)
library(viridis)
library(dplyr)
```

Lets use some station data from water temperature dataset.
```{r, warning = FALSE, message = FALSE}
stations_URL <- "https://portal.edirepository.org/nis/dataviewer?packageid=edi.591.2&entityid=a059f5eea4f8500fe1a43566451ec593"

stations <- readr::read_csv(stations_URL) %>%
  mutate(County = factor(County)) %>%
  mutate(Temperature = runif(n = nrow(.), min = 8, max = 30)) # Add a column of made up temperature data
dplyr::glimpse(stations)
```

Define your color palette. How do you want to color your stations?
```{r, warning = FALSE, message = FALSE}
# Let's say we are coloring by County. 
unique(stations$County) # There are 6 counties

# We will pick 6 colors from the viridis package. 
pal <- colorFactor(viridis::viridis(6), stations$County)
```

You could alternatively use colorNumeric, colorBin, or colorQuantile as an argument. See https://rstudio.github.io/leaflet/colors.html

Make your plot. 
```{r, warning = FALSE, message = FALSE}
stations %>% # call data frame.
  leaflet() %>% # call leaflet.
  addTiles() %>% # this adds the map in the background.
  addCircleMarkers(
    color = ~pal(County),
    stroke = FALSE, # alters whether there is a border to circle
    fillOpacity = 0.8,
    lng = ~Longitude, # call your longitude column name
    lat = ~Latitude, # call you latitude column name
  label = ~paste(Station, "Lat:", Latitude, "Long:", Longitude)) %>% # edit what you want to show up in your label
  addLegend(pal = pal,
            values = ~County,
            position = "bottomright")
```

If you want to change the radius of your points based on a variable, you can also do that. 

Define your palette colors again. In this example, we picked individual colors rather than using an existing palette, and assign them directly to each basin.
```{r, warning = FALSE, message = FALSE}
unique(stations$Basin) # need 7

pal2 <- colorFactor(c("purple3", "blue3", "forestgreen", "orange", "orangered", "hotpink", "grey"), domain = c("Cache Cr", "Delta", "East Bay", "Mokelumne R", "Sacramento R", "San Joaquin R", "Stanislaus R"))
```

Make map, this time defining the "radius" option. 
```{r, warning = FALSE, message = FALSE}
stations %>%
  leaflet() %>%
  addTiles() %>%
  addCircleMarkers(
    fillColor = ~pal2(Basin),
    radius = ~sqrt(Temperature), # modify the size of circles based on the temperature
    stroke = FALSE, 
    fillOpacity = 0.6,
    lng = ~Longitude,
    lat = ~Latitude,
    label = ~paste(Station, "Lat:", Latitude, "Long:", Longitude, "Temperature:", Temperature))%>%
  addLegend(pal = pal2,
            values = ~Basin,
            position = "topright")
```

You can also have the labels be more permanent.
See labelOptions below (noHide = T). If textOnly = FALSE, there will be a white box around your label. Doesn't work great for this plot, since there are so many stations. 
```{r, warning = FALSE, message = FALSE}
stations %>%
  leaflet() %>%
  addTiles() %>%
  addCircleMarkers(
    fillColor = ~pal2(Basin),
    radius = ~sqrt(Temperature), # modify the size of circles based on the temperature
    stroke = FALSE, 
    fillOpacity = 0.6,
    lng = ~Longitude,
    lat = ~Latitude,
    label = ~Station,
    labelOptions = labelOptions(noHide = T, direction = "right", textsize = "7px", textOnly = TRUE))%>%
  addLegend(pal = pal2,
            values = ~Basin,
            position = "topright")
```

There are other types of markers too.
This is the default.
```{r, warning = FALSE, message = FALSE}
stations %>%
  leaflet() %>%
  addTiles() %>%
  addMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    label = ~Station)
```

This is a square. 
```{r, warning = FALSE, message = FALSE}
stations %>%
leaflet() %>% 
  addTiles() %>%
  addRectangles(
    lng1=min(stations$Longitude) - 0.1, lat1=min(stations$Latitude) -0.1 ,
    lng2=max(stations$Longitude) +0.1, lat2=max(stations$Latitude) +0.1,
    fillColor = "transparent",
    color = "orangered"
  ) %>%
  addCircleMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    label = ~Station,
    radius = 2)
```

## Mapview

This one is a nice, simple way to just take a look at your points.
```{r, warning = FALSE, message = FALSE}
library(mapview)
library(sf)
```

Convert your stations to a spatial object. 
```{r, warning = FALSE, message = FALSE}
# name the columns for longitude and latitude
# crs is the WGS 1984 projection, which is usually the projection for lat/lon data
stations_sf <- sf::st_as_sf(stations, coords = c("Longitude", "Latitude"), crs = 4326)
```

Just one line to see a decent plot!
This one shows you all the associated data linked with each point.
```{r, warning = FALSE, message = FALSE}
mapview(stations_sf)
```

Color by variable of interest
```{r, warning = FALSE, message = FALSE}
mapview(stations_sf, zcol = "Basin")
```

Size your variables
```{r, warning = FALSE, message = FALSE}
mapview(stations_sf, zcol = "Basin", cex = "Temperature")
```
