---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Spatial data and making maps

We will cover:
1. Basics of using sf package, converting lat/lon to a spatial data frame.
2. Importing shapefiles
3. Making a map of your lats/lons
4. Add inset map
5. Add scale bar, north arrow, labels
6. Add basemap
7. Writing map image and map files 


Packages
```{r}
library(ggplot2)
library(dplyr)
library(sf)
library(viridis)
```

## Load data

Latitude/Longitude data from IEP zooplankton data
```{r}
# Latitude/Longitude data
stations_URL <- "https://portal.edirepository.org/nis/dataviewer?packageid=edi.539.3&entityid=343cb43b41eb112ac36b605f1cff1f92"

# Create a fake n variable to have something to plot.
stations <- readr::read_csv(stations_URL) %>%
  mutate(n = round(runif(n = 1:nrow(.), min = 1, max = 100),0)) %>%
  mutate(Source = factor(Source)) %>%
  filter(!is.na(Latitude))

dplyr::glimpse(stations)

# create a subset of stations
stationlist_filt <- sample_n(stations, 20) %>% select(Station)
stationlist_filt <- stationlist_filt$Station
```

Shapefiles
```{r}
# Delta waterways
library(deltamapr)
plot(WW_Delta)
glimpse(WW_Delta)

# Regions
Regions <- deltamapr::R_EDSM_Subregions_Mahardja_FLOAT
plot(Regions)
glimpse(Regions)

# States
library(USAboundaries)
California_sf <- us_states(states = "California")
plot(California_sf[1])
```

## Get data into spatial form
```{r}
# Define the projection of your points, usually WGS 84 (= crs 4326)
stations_sf <- st_as_sf(stations, coords = c("Longitude", "Latitude"), crs = 4326)

# Look at shapefile
head(stations_sf)
plot(stations_sf$geometry)
plot(stations_sf)
```


### Spatial projections

You want to make sure all your different files are in the same projection, or they will look mis-aligned.
```{r}
st_crs(WW_Delta) # In 4269
st_crs(stations_sf) # In 4326
st_crs(California_sf) # In 4326

stations_4269 <- st_transform(stations_sf, crs = 4269)
california_4269 <- st_transform(California_sf, crs = 4269)
```

## Basic Spatial Operations

Assign points to regions
```{r}

```

Nearest neighbor
```{r}

```

Intersections
```{r}

```


## Make maps

### Basic maps

Color by Monitoring Program
```{r}
ggplot() +
    geom_sf(data = WW_Delta) +
    geom_sf(data = stations_4269, aes(fill = Source), shape = 21) + 
    scale_fill_viridis(discrete = TRUE) + 
    ggtitle("Number of Zooplankton Samples by Station") +
    theme_bw()
```

You can also modify the size of the points by size
```{r}
# Making a smaller dataset
stations_filtered_4269 <- stations_4269 %>%
  filter(Station %in% stationlist_filt)

(simplemap <- ggplot() +
    geom_sf(data = WW_Delta, fill = "lightblue", color = "lightblue") +
    geom_sf(data = stations_filtered_4269, aes(fill = Source, size = n), shape = 21, alpha = 0.7) + 
    scale_fill_viridis(discrete = TRUE, option = "plasma") + 
    scale_size_continuous(range = c(0,3)) + 
    ggtitle("Number of Zooplankton Samples by Station") +
    theme_bw())
```

You can add region delineations as well
```{r}
ggplot() +
    geom_sf(data = WW_Delta, fill = "lightblue", color = "lightblue") +
    geom_sf(data = Regions, aes(color = Region), alpha = 0.3) +
    geom_sf(data = stations_filtered_4269, aes(fill = Source, size = n), shape = 21) + 
    scale_fill_viridis(discrete = TRUE, option = "plasma") + 
  scale_color_viridis(discrete = TRUE, option = "turbo") +
    scale_size_continuous(range = c(0,3)) + 
    ggtitle("Number of Zooplankton Samples by Station") +
    theme_bw()
```


### Add arrows and scale bars, dotted lines
```{r}
require(ggspatial)
# https://www.r-spatial.org/r/2018/10/25/ggplot2-sf.html

(simplemap2 <- simplemap + 
  annotation_north_arrow(location = "tr", which_north = "true", 
        pad_x = unit(0.1, "in"), pad_y = unit(0.1, "in"),
        style = north_arrow_fancy_orienteering) +
  annotation_scale(location = "bl", bar_cols = c("pink", "white", "pink", "white")) +
  theme(axis.title = element_blank(),
        panel.grid.major = element_line(color = "grey80", linetype = "dashed", size = 0.5)))
```

### Add labels to map
```{r}
# Adding text
simplemap2 + 
  geom_text(data = filter(stations, Station %in% stationlist_filt), aes(x = Longitude, y = Latitude, label = Station), size = 2, check_overlap = FALSE, color = "darkblue", nudge_x = 0.02, nudge_y = 0.02) + 
 annotate(geom = "text", x = -122.4, y = 37.85, label = "San Francisco Bay", fontface = "italic", color = "grey22", size = 3.5 ) 
```

### Add inset map
```{r}
# Figure out boundary box for your stations; perhaps add a small buffer
insetbbox0 = st_as_sfc(st_bbox(WW_Delta))
insetbbox = st_buffer(insetbbox0, 0.2)

(inset <- ggplot() + 
  geom_sf(data = california_4269, fill = "white") +
  geom_sf(data = insetbbox0, fill = NA, color = "red", size = 0.5) +
  theme_void())
```

Combine main map with inset map

Will need to play with where you want the inset to be so as not to overlap with your map
```{r}
library(cowplot)
inset_map = ggdraw() +
  draw_plot(simplemap2) +
  draw_plot(inset, x = 0.15, y = 0.63, width = 0.3, height = 0.3)

inset_map
```

## Basemaps
Download basemaps from get_stamenmap
```{r}
library(ggmap)
```

Define coordinate bounding box. You could also use lat/lon if you want.
```{r}
buffer = 0.2
coordDict = list( 
    'minLat' = min(stations$Latitude) - buffer,
    'maxLat' = max(stations$Latitude) -0.1,
    'minLon' = min(stations$Longitude) - buffer,
    'maxLon' = max(stations$Longitude) + buffer
)

# Create map object using your bounded coordinates
map_obj <- get_stamenmap(
  bbox = c(left = coordDict[['minLon']], bottom = coordDict[['minLat']], right = coordDict[['maxLon']], top = coordDict[['maxLat']]), # the bounding box
  zoom = 9, # zoom lvl; higher number = more detail (but also more processing power)
  maptype = 'terrain-background'# type of basemap; 'terrain' is my default, but check help(get_stamenmap) for a full list
  )
```

Plot your basemap
```{r}
# Plot the map
map <- ggmap(map_obj, legend = "right")
map
```

Add basemap to earlier map.
```{r}
map2 <- ggmap(map_obj) +
    geom_sf(data = WW_Delta, fill = "lightblue", color = "lightblue", inherit.aes = FALSE) + 
    geom_sf(data = stations_filtered_4269, aes(fill = Source), shape = 21, alpha = 0.7, size = 2.5, inherit.aes = FALSE) + 
 annotate(geom = "text", x = -122.4, y = 37.85, label = "San Francisco Bay", fontface = "italic", color = "grey22", size = 3.5 ) +
  annotation_north_arrow(location = "tr", which_north = "true", 
        pad_x = unit(0.1, "in"), pad_y = unit(0.1, "in"),
        style = north_arrow_fancy_orienteering) +
  annotation_scale(location = "bl", bar_cols = c("black", "white", "black", "white")) +
    scale_fill_viridis(discrete = TRUE, option = "plasma") + 
    ggtitle("Number of Zooplankton Samples by Station") +
    theme_bw()+ theme(axis.title = element_blank(),
        panel.grid.major = element_line(color = "grey80", linetype = "dashed", size = 0.5))
map2
```

